---
title: "Using promiscuous drugs to reveal novel therapeutic targets in tauopathy"
gh-repo: shkoren/Comp-bio-projects/Tau-Ribosome Association
subtitle: Coupling broad kinase inhibition with quantitative TMT proteomics
tags:
- tau
- proteomics
- R
thumbnail-img: /assets/img/kinase.png
gh-badge:
- star
- watch
- follow


---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi=200, fig.width=6,fig.height = 4)
knitr::opts_knit$set(root.dir = here::here())


```

This code demonstrates the proteomic analysis used in our recent preprint, "Broad kinase inhibition mitigates early neuronal dysfunction and cognitive deficits in tauopathy" found [here](https://www.biorxiv.org/content/10.1101/2020.07.31.229583v1). Before delving into the code and figures behind the proteomics, I'll provide some background on the study.

## Introduction

<!-- Kinase inhibitor drugs provide powerful therapeutic potential in the treatment of chronic diseases like cancer, inflammation, and neurodegenerative diseases. Nearly 50 kinase inhibitors have been approved for clinical use today, and hundreds more are currently undergoing evaluation in clinical trials. The target promiscuity of these drugs, or their off-target effects, are often under-reported or under-invesetigated. This leaves the entirety of the drug's effect hidden from view until patient symptoms arise. Moreover, some drugs retain putative therapeutic aspects even after their canonical protein target is knocked-out, suggesting the off-target effects are powering any disease modifying effects  [Klaeger, Science, 2017](https://science.sciencemag.org/content/358/6367/eaan4368); [Lin, STM, 2019](https://stm.sciencemag.org/content/11/509/eaaw8412)). When we ascribe disease-modifying qualities to drugs whose target is unknown, we distort our understanding of disease and of biology. -->

<!-- During this study's preparation, the kinase inhibitor used (GSK2606414, or 414) was found to have considerable disease-relevant targets beyond its canonical substrate, PKR-like Endoplasmic Reticulum Kinase, or PERK. 414 had been used in, and indeed reported therapeutic potential for, dozens of studies across numerous diseases and tissues. To date, 414 inhibits over 20 kinases at IC50 < 4 μM, and nearly 75 at <10 μM  [ref]. During the last year of preparation, I argued to reshape the narrative of the paper to directly focus on the promiscuity of 414 and use quantitiative proteomics to analyze its effects to gleam biological insights on tauopathy. -->

<!-- between mouse models and human tauopathy We used early stage tauopathy model mice which exhibit the initial signs of Alzheimer's disease (AD) and other tauopathies: neurodegeneration, cognitive dysfunction, and pathological tau inclusions. As this early stage has the greatest therapeutic potential for the treatment of of e We used a tauopathy mouse model at an age presenting early signs of tauopathic dementia: brain atrophy, progressive cognitive decline, and accumulating inclusions of pathological tau.  Originally, 414 was used to delay the onset of the integrated stress response (ISR) via blocking the ER stress kinase PERK from phosphorylation eukaryotic initiation factora 2 alpha, or eiF2α. However, over the course of the study, GSK2606414 As the onset of PERK activity and resulting ISR occurs at an older age than the one used in this study, we suspected the drug would have limited effects until the outcome of delaying ISR activity became apparent [ref]. Instead, our early magnetic resonance imaging revealed the drug ameliorated brain atrophy and neuronal calcium dysfunction, outcomes previously thought to correspond to 414-mediated PERK inhibition. The resulting study was formed to investigate the molecular sources of that cognitive rescue, in particular via LC-MS/MS hippocampal proteomics. -->

Tauopathies are a group of over 20 neurodegenerative diseases involving progressive brain atrophy, cognitive dysfunction, and accumulation of pathological tau protein with no disease modifying treatments available. This lack of therapeutic strategies is partly due to poor understanding of the molecular mechanisms underlying these complex disease features. Dysregulated kinase activity has consistently been shown to be an early hallmark of tauopathies. Aberrant kinase signaling directly causes maladaptive signaling cascades which disrupt vital cellular processes and induce tau hyper-phosphorylation and subsequent toxicity.

Here, we investigated whether broad kinase inhibition rescues the cognitive and molecular outcomes of tauopathy by using a mouse model of the primary tauopathy, frontotemporal dementia (FTD). Treatment with GSK2606414 (414) ameliorated brain atrophy, cognitive and behavioral pheonotypes, and measures of neuronal calcium function. As 414 targets over 20 kinases at IC50 < 4 μM and nearly 75 at <10 μM, we evaluated the hippocampal proteomic signatures of drug treatment in tau transgenic mice to reveal:
1. The altered hippocampal proteome in FTD model mice relative to control, non-transgenic mice,
2. The effects of broad kinase inhibition associated with cognitive rescue, and
3. High confidence candidate targets for therapeutic intervention in human tauopathy


## TMT pathway initial steps

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE, fig.align = 'center'}
#Initializing data and packages
library(dplyr)
# library(broom)
library(tidyverse)
# library(ComplexHeatmap)
# library(reshape2)
library(ggplot2)
# library(RColorBrewer)
library(cowplot)
library(here)
library(janitor)
library(qdapRegex)

#knitr::opts_knit$set(root.dir = here::here())

options(scipen = 999) #Turn off sci-notation

rawDataFile = 'TMT_raw.txt'
rawData = read.delim(here("data", rawDataFile), fileEncoding="UTF-8-BOM")
rawDataClean = rawData %>% clean_names() %>% filter(master == 'IsMasterProtein') %>% filter(x_unique_peptides > 0) %>% select(c(6:7, 44:64, 194:215)) %>% na.omit() %>%  mutate(Symbol = unlist(rm_between(description, "GN=", " PE", extract=TRUE))) %>% select(Symbol, everything()) %>% rename(., Uniprot = accession)

test = rawData %>% select(194:215)

protMatNorm = rawDataClean %>% select(c(1:3, 4:23)) %>% gather(Sample, NormalizedAbundance, 4:23) %>% mutate(Batch = substr(Sample, 18, 18)) %>%
  mutate(Tag = unlist(rm_between(Sample, c("ratio_f1_","ratio_f2_"), c("_f1_131","_f2_131"), extract=TRUE))) 
protMat = rawDataClean %>% select(c(1:3, 25:46)) %>% gather(Sample, Abundance, 4:25) %>% mutate(Batch = substr(Sample, 12, 12)) %>%
  mutate(Tag = substr(Sample, 14, 17)) %>% mutate(Tag = case_when(Tag == "126_" ~ "126", TRUE ~ as.character(Tag)))

p2 = ggplot(data = protMatNorm, aes(x = toupper(Tag), y = NormalizedAbundance, fill = Batch)) + geom_boxplot() + 
  theme(axis.text.x = element_text(face = "bold", angle = 45, hjust = 1)) + xlab("TMT Tag") + ylab("Standard (131C) Normalized Intensity")
p1 = ggplot(data = protMat, aes(x = toupper(Tag), y = log2(Abundance), fill = Batch)) + geom_boxplot() + 
  theme(axis.text.x = element_text(face = "bold", angle = 45, hjust = 1)) + xlab("TMT Tag")

plot_grid(p1, p2, ncol = 2, labels = "AUTO")

```